\section{Identifiability and agreement consistency}
\label{sec:identifiability}

We port \cref{thm:bg-id} to the label model. Three results follow:
the glass ceiling (T1), identifiability under conditional
independence (T2), and agreement consistency (T3).

\subsection{The glass ceiling: gold-free non-identifiability}

The label model is fit from LF votes alone. Whether its parameters
can be recovered from the vote distribution is an identifiability
question, and the answer without further structure is negative.

\begin{theorem}[Glass ceiling, gold-free non-identifiability]
\label{thm:glass-ceiling}
Let $\bm{\lambda}_i$ be the LF votes under the DGP of
\cref{sec:translation}, with unknown accuracies $\bm{\alpha}$,
coverages $\bm{\beta}$, and class prior $\pi$. Without gold-labeled
examples, and absent a structural assumption on LF dependence or LF
orientation, the pair $(\bm{\alpha}, \pi)$ is non-identifiable from
the distribution of LF votes: there exist distinct parameter sets
producing identical vote distributions. In particular the
accuracy-complement symmetry holds: replacing $(\pi, \bm{\alpha})$ by
$(1 - \pi,\ \1 - \bm{\alpha})$ leaves every joint vote probability
unchanged.
\end{theorem}

\begin{proof}[Sketch]
We exhibit non-identifiability by construction. Fix any example with
votes $\bm{\lambda}$. Under the symmetric accuracy model, the joint
probability of the votes marginalizing over $Y$ is
\begin{equation}
  \Prob\{\bm{\lambda}\}
  = \sum_{y \in \{0,1\}} \Prob\{Y = y\}
    \prod_{j:\,\lambda_j \neq \varnothing}
    \beta_j\, \alpha_j^{\1\{\lambda_j = y\}}
    (1-\alpha_j)^{\1\{\lambda_j \neq y\}}
    \prod_{j:\,\lambda_j = \varnothing} (1 - \beta_j).
  \label{eq:id-jointvote}
\end{equation}
Apply the relabeling $y \mapsto 1 - y$ inside the sum. The prior term
$\Prob\{Y = y\}$ becomes $\Prob\{Y = 1 - y\}$, which is the prior of
the parameter set with $\pi$ replaced by $1 - \pi$. Each factor
$\alpha_j^{\1\{\lambda_j = y\}}(1-\alpha_j)^{\1\{\lambda_j \neq y\}}$
becomes $\alpha_j^{\1\{\lambda_j \neq y\}}(1-\alpha_j)^{\1\{\lambda_j
= y\}}$, which is the factor of the parameter set with $\alpha_j$
replaced by $1 - \alpha_j$. The coverage factors are untouched. Hence
$(\pi, \bm{\alpha})$ and $(1 - \pi, \1 - \bm{\alpha})$ produce
identical $\Prob\{\bm{\lambda}\}$ for every vote pattern, so they are
observationally indistinguishable. An LF of accuracy $\alpha_j > 1/2$
is statistically a mirror image of one with accuracy $1 - \alpha_j <
1/2$ that has its output label flipped. The two solutions are
separated only by an external assumption: ``every LF is better than
random'' ($\alpha_j > 1/2$, fixing each LF's orientation), a fixed
class prior, or conditional independence with a known sign on one
agreement moment. The binary case exhibits the $\mathbb{Z}/2$
label-permutation element of the symmetric group; the multiclass
analog is the full symmetric group $\mathfrak{S}_K$ acting on
$\{0, 1, \ldots, K-1\}$, and the corresponding obstruction is
non-identifiability up to a label permutation. The reduction of the
gold-free obstruction to a rank condition on the augmented
candidate-set matrix is \cref{thm:bg-id} of \cref{sec:background};
the full apparatus, including regularity conditions, is in
\citet{towell2026masked}.
\end{proof}

\Cref{thm:glass-ceiling} is the weak-supervision instance of the
glass-ceiling phenomenon: an estimator that uses only LF votes cannot
orient the label model. It is the same structural obstruction that
\citet{towell2026scrnacoarsening} identify for scRNA-seq zeros, here
made fully explicit as a label-flip symmetry. The same obstruction
appears in the partial-labeler model of \citet{yu2022nplm}, who prove
gold-free identifiability only up to label swapping; \cref{eq:id-jointvote}
exhibits the binary witness of that swap symmetry in closed form and
ties the residual ambiguity to a rank deficit of the augmented
candidate-set matrix rather than leaving it as a generic permutation
caveat. A single gold-labeled example breaks the symmetry: it is a
singleton candidate set whose known label fixes the orientation,
adding an identifying row to the augmented candidate-set matrix of
\cref{thm:bg-id}. The
better-than-random assumption is the cost-free alternative when it is
credible; the next theorem shows what conditional independence buys
without any gold data.

\subsection{Identifiability under conditional independence}

The standard rescue in data programming is the assumption that LFs
err independently given the label. Under that assumption the label
model is identifiable from agreement moments alone.

\begin{theorem}[Identifiability under conditional independence]
\label{thm:ci-identifiability}
Suppose the LF votes are conditionally independent given $Y$, that at
least three LFs have non-degenerate accuracy ($\alpha_j \neq 1/2$ and
$0 < \beta_j \le 1$), and that LF orientations are fixed (e.g., every
LF is better than random). Then the accuracies $\bm{\alpha}$ and the
class prior $\pi$ are identifiable from the second- and third-order
LF agreement moments.
\end{theorem}

\begin{proof}[Sketch]
Encode each non-abstaining vote as $\sigma_{ij} = 2\,\1\{\lambda_{ij}
= 1\} - 1 \in \{-1, +1\}$ and the latent label as $Z = 2Y - 1$.
Conditional independence gives, for distinct LFs $j, k, \ell$ that
all vote,
\begin{equation}
  \E[\sigma_j \sigma_k] = a_j a_k, \qquad
  \E[\sigma_j \sigma_k \sigma_\ell] = a_j a_k a_\ell,
  \label{eq:id-triplet}
\end{equation}
where $a_j = \E[\sigma_j Z] = 2\alpha_j - 1$ is the LF's
\emph{accuracy margin} (its centered accuracy). Equation
\eqref{eq:id-triplet} is the moment structure exploited by the
triplet method of \citet{fu2020fast}. From three LFs, the three
pairwise products and the one triple product give
\begin{equation}
  a_j^2 = \frac{\E[\sigma_j \sigma_k]\,\E[\sigma_j \sigma_\ell]}
               {\E[\sigma_k \sigma_\ell]},
  \label{eq:id-solve}
\end{equation}
and symmetric expressions for $a_k, a_\ell$; the triple product fixes
the common sign, and the fixed-orientation assumption (or the
better-than-random assumption) selects $a_j > 0$. Each $a_j$
determines $\alpha_j = (1 + a_j)/2$. With the margins known, the
first moment $\E[\sigma_j] = a_j\,\E[Z] = a_j(2\pi - 1)$ identifies
$\pi$. Non-degeneracy ($a_j \neq 0$) keeps every denominator nonzero.
This is the data-programming and Dawid--Skene identifiability result
\citep{dawid1979maximum,ratner2016data,fu2020fast} re-derived as a
masked-cause statement: conditional independence makes the augmented
candidate-set matrix full column rank, so \cref{thm:bg-id} applies
and $(\bm{\alpha}, \pi)$ is identifiable without gold data.
\end{proof}

\Cref{thm:ci-identifiability} is the positive counterpart of T1.
Conditional independence is the structural assumption that breaks the
glass ceiling without gold labels: it supplies the moment equations
\eqref{eq:id-triplet} whose rank is exactly the augmented
candidate-set rank of \cref{thm:bg-id}. The fixed-orientation
assumption (each LF is declared to vote on the positive side of its
target class) is a checkable property of each LF's source code; it
is not automatic, however, since an LF written backward (a regex
that positively matches the negative class) reverses its sign and
silently violates the assumption. The assumption is also the
fragile one. When LFs share error structure (two regex LFs keyed on
the same token, two model-based LFs sharing a backbone), the products
in \eqref{eq:id-triplet} pick up cross terms, the rank drops, and the
gold-free estimator is biased. \Cref{sec:methodology} quantifies how
much gold data restores the lost rank.

\subsection{Agreement consistency}

A structural property holds for any label model fit by maximum
likelihood, regardless of whether the conditional-independence
assumption is correct: the fitted model reproduces the empirical
pairwise LF agreement rates.

\begin{theorem}[Agreement consistency, regular exponential family]
\label{thm:agreement-consistency}
Let $(\hat{\bm{\alpha}}, \hat{\pi})$ be a maximum-likelihood estimate
of the label model for a parametrization that forms a regular
exponential family whose sufficient statistics include the pairwise
LF agreement indicators $\1\{\lambda_{ij} = \lambda_{ik}\}$ (the
standard naive-Bayes parametrization of \citet{ratner2016data} does
not satisfy this hypothesis exactly; an asymptotic version of the
identity for the naive-Bayes corner is given in
\cref{cor:agreement-nb} below). Then for
every pair of LFs $(j, k)$,
\begin{equation}
  \widehat{\Prob}\{\lambda_j = \lambda_k \mid \text{both vote}\}
  = \bar{A}_{jk},
  \label{eq:agreement}
\end{equation}
where $\bar{A}_{jk}$ is the empirical agreement rate of LFs $j$ and
$k$ on the examples where both vote. The identity holds exactly at
any \emph{interior} MLE (no $\hat\alpha_j \in \{0,1\}$, no
$\hat\beta_j \in \{0,1\}$, $\hat\pi \in (0,1)$); boundary cases
reduce to degenerate-LF identities.
\end{theorem}

\begin{proof}[Sketch]
The argument parallels the cell-total consistency proof of
\citet[\S 3]{towell2026scrnacoarsening}. For a regular exponential
family the MLE satisfies the moment-matching identity
$\E_{\hat\theta}[T(\bm{\lambda})] = \bar{T}$ for every sufficient
statistic $T$ (the stationarity of the log-partition gradient). When
$T$ is the pair-$(j,k)$ agreement indicator
$\1\{\lambda_{ij} = \lambda_{ik}\}$ this is exactly
\eqref{eq:agreement}. The full derivation, including the boundary
regimes, is the moment-matching argument of
\citet{towell2026scrnacoarsening} with the candidate-set sufficient
statistic specialized to LF agreement.
\end{proof}

The conditioning matters: the identity \eqref{eq:agreement} is a
sufficient-statistic consequence, so it follows whenever the
parametrization is rich enough to count pairwise agreement among its
sufficient statistics. The standard naive-Bayes label model of data
programming \citep{ratner2016data} parametrizes by per-LF accuracies
$\bm{\alpha}$ and the class prior $\pi$; the pairwise agreement
indicators are \emph{not} per-pair sufficient statistics of that
parametrization. The result still applies in a weaker, asymptotic
sense for the conditionally independent corner:

\begin{corollary}[Asymptotic agreement consistency for naive-Bayes]
\label{cor:agreement-nb}
Under the conditional-independence assumption of
\cref{thm:ci-identifiability}, the naive-Bayes label model
parametrized by $(\bm{\alpha}, \pi)$ satisfies
\eqref{eq:agreement} asymptotically: at the population
parameter values the joint vote factorization implies $\Prob\{
\lambda_j = \lambda_k\} = \alpha_j \alpha_k + (1-\alpha_j)(1-\alpha_k)$
exactly, and the MLE converges to that value at the standard
$n^{-1/2}$ rate. The identity holds in the limit, not at every finite
sample.
\end{corollary}

The simulation in \cref{sec:validation} confirms this asymptotic
behavior at large $n$: the agreement residual decays as $n^{-1/2}$
and reaches Monte-Carlo noise at $n = 500{,}000$. The simulation is
consistent with \cref{thm:agreement-consistency} for the extended
exponential-family parametrization and with \cref{cor:agreement-nb}
for the naive-Bayes parametrization; it does not isolate the
sufficient-statistic property structurally.

\Cref{thm:agreement-consistency} is the label-model analogue of
cell-total consistency. Its practical force is a warning. Two label
models, one assuming conditional independence and one fit with a
dependency graph, can both reproduce every observed pairwise
agreement rate exactly (the first via \cref{cor:agreement-nb}
asymptotically, the second via \cref{thm:agreement-consistency} at
the exact MLE), while disagreeing on the LF accuracies and hence on
the probabilistic training labels. A practitioner checking ``does
the fitted model match the observed LF agreement?'' will see a
perfect fit from a biased model. Diagnosing accuracy bias requires
either gold labels or an external structural assumption; the marginal
agreement fit cannot do it. The same caution was raised for the ZINB
MLE in \citet{towell2026scrnacoarsening}; here it is sharper, because
the dependent and independent label models satisfy
\eqref{eq:agreement} jointly yet imply different supervision.
Agreement consistency (\cref{thm:agreement-consistency}) is a special
case of the general coarsened-data consistency theorem of
\citet{towell2026synthesis}, in which the singleton-candidate-set
device used here, gold labels, appears as one instance of a
domain-independent rank condition and singleton-restoration mechanism.
